---
title: "Performing Two-Sample Mean Comparison Using HMC"
author: "Tianyu Zhang"
format: html
editor: visual
---

This tutorial demonstrates how to use the `HMC` package to perform high-dimensional two-sample mean comparison, leveraging data-driven projection and cross-fitting for improved power and interpretability. The demonstration uses a sample dataset that can be downloaded via the following Bash script:

```{bash, eval = FALSE}
wget https://github.com/terrytianyuzhang/HMC/blob/main/docs/data/tiny_cleary_for_HMC.rds
```

```{r, echo = TRUE, eval=FALSE}
library(data.table)
library(HMC)

# Load data
residual_subset <- readRDS("tiny_cleary_for_HMC.rds")

# Subset to control and treatment groups, and drop guide annotations

control <- residual_subset[
    Guides_collapsed_by_gene == "non-targeting",
    !"Guides_collapsed_by_gene"
]
perturbed <- residual_subset[
    Guides_collapsed_by_gene == "STAT1",
    !"Guides_collapsed_by_gene"
]

```

```{r, echo = FALSE, eval=TRUE}
library(data.table)
library(HMC)

# Load data
output_dir <- "/Users/tianyuzhang/Documents/HMC/docs/"
residual_subset <- readRDS(paste0(output_dir, "/data/tiny_cleary_for_HMC.rds"))
clustering <- readRDS(paste0(output_dir, "/data/gene_clustering_for_HMC.rds"))

# Subset to control and treatment groups, then remove guide labels
control <- residual_subset[
    Guides_collapsed_by_gene == "non-targeting",
    !"Guides_collapsed_by_gene"
]
perturbed <- residual_subset[
    Guides_collapsed_by_gene == "STAT1",
    !"Guides_collapsed_by_gene"
]
```

First 3 cells and first 5 genes from the control group:

```{r, eval=TRUE}
control[1:3, 1:5]
```

First 3 cells and first 5 genes from the treatment group:

```{r, eval = TRUE}
perturbed[1:3, 1:5]
```

Next, we apply the test using a subset of genes corresponding to a specific cluster. We load the gene grouping file and isolate genes from cluster 31:

```{r, echo = FALSE, eval = TRUE}
clustering <- readRDS(paste0(output_dir, "/data/gene_clustering_for_HMC.rds"))
gene_to_keep <- which(clustering$cluster_index == 31)
control_subset <- control[, ..gene_to_keep]
perturbed_subset <- perturbed[, ..gene_to_keep]
```

```{r, echo = TRUE, eval = FALSE}
clustering <- readRDS("gene_clustering_for_HMC.rds")
gene_to_keep <- which(clustering$cluster_index == 31)
control_subset <- control[, ..gene_to_keep]
perturbed_subset <- perturbed[, ..gene_to_keep]
```

We now run the comparison function using the selected genes:

```{r, echo = TRUE, eval = FALSE}
set.seed(123)
test_result <- mean_comparison_anchor(
    control = control_subset,
    treatment = perturbed_subset,
    pca_method = "sparse_pca",
    classifier_method = "lasso",
    lambda_type = "lambda.min",
    n_folds = 5,
    verbose = FALSE
)
```

```{r, echo = FALSE, eval = FALSE}
saveRDS(test_result, file = paste0(output_dir, "data/lasso_group31_test_result.rds"))
```

```{r, echo = FALSE, eval = TRUE}
test_result <- readRDS(paste0(output_dir, "data/lasso_group31_test_result.rds"))
```

The p-value from the two-sample mean comparison test is:

```{r, eval = TRUE}
test_result$p_value
```

The function below extracts the genes most frequently selected by the lasso classifier as discriminative features between the two groups:

```{r, eval = TRUE}
collect_active_features_proj(test_result)
```