---
title: "Performing Convergence Analysis Using HMC"
format:
  html:
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
editor: visual
---

This tutorial demonstrates how to perform convergence analysis using the `HMC` package. We will use a sample dataset, which can be downloaded using the following command.

<!-- You can run it directly in your terminal and the data will, normally, be downloaded in the folder where you ran it. -->

````{=html}
<!-- ```{bash, eval = FALSE}
wget https://github.com/terrytianyuzhang/HMC/blob/main/docs/data/tiny_cleary_for_HMC.rds
``` -->
````

```{r, eval = FALSE}
my_local_path <- "~/your/local/path/"
url <- "https://raw.githubusercontent.com/terrytianyuzhang/HMC/main/docs/data/tiny_cleary_for_HMC.rds"
dest_file <- paste0(my_local_path, "tiny_cleary_for_HMC.rds")
download.file(url, dest_file, mode = "wb") 
```

To get started, install the `HMC` package from GitHub using the following command:

```{r, eval = FALSE}
devtools::install_github("terrytianyuzhang/HMC/HMC_package")
```

Next, load the required packages and data, then extract the control and treatment groups for the convergence analysis. We will use `STAT1` and `STAT2`. Based on their names, we expect them to induce similar effects when perturbed.

```{r, echo = TRUE, eval=FALSE}
library(data.table)
library(HMC)

# Load data
residual_subset <- readRDS(paste0(my_local_path, "tiny_cleary_for_HMC.rds"))

# Subset to control and treatment groups, then remove guide labels
control <- residual_subset[
  Guides_collapsed_by_gene == "non-targeting",
  !"Guides_collapsed_by_gene"
]
STAT1 <- residual_subset[
  Guides_collapsed_by_gene == "STAT1",
  !"Guides_collapsed_by_gene"
]

STAT2 <- residual_subset[
  Guides_collapsed_by_gene == "STAT2",
  !"Guides_collapsed_by_gene"
]
```

```{r, echo = FALSE, eval= TRUE}
library(data.table)
library(HMC)

# Load data
output_dir <- "~/Documents/HMC/docs/"
residual_subset <- readRDS(paste0(output_dir, "/data/tiny_cleary_for_HMC.rds"))

# Subset to control and treatment groups, then remove guide labels
control <- residual_subset[
  Guides_collapsed_by_gene == "non-targeting",
  !"Guides_collapsed_by_gene"
]
STAT1 <- residual_subset[
  Guides_collapsed_by_gene == "STAT1",
  !"Guides_collapsed_by_gene"
]

STAT2 <- residual_subset[
  Guides_collapsed_by_gene == "STAT2",
  !"Guides_collapsed_by_gene"
]
```

First 3 cells and first 5 genes from the control group:

```{r, eval=TRUE}
control[1:3, 1:5]
```

First 3 cells and first 5 genes from the STAT1 group:

```{r, eval = TRUE}
STAT1[1:3, 1:5]
```

## 1. Gene-Level Perturbation Convergence Analysis

Now we perform the convergence analysis using the `convergence_testing()` function.

We compare `STAT1` and `STAT2`, using control group as the baseline.

```{r, echo = TRUE, eval = FALSE}
set.seed(123)
test_result <- convergence_testing(
  control = control,
  treatment1 = STAT1,
  treatment2 = STAT2,
  pca_method = "dense_pca",
  classifier_method = "lasso",
  lambda_type = "lambda.min",
  n_folds = 5,
  verbose = TRUE
)
```

```{r, echo = FALSE, eval = FALSE}
saveRDS(test_result, file = paste0(output_dir, "data/HMC_convergence_control_STAT1_STAT2.rds"))
```

```{r, echo = FALSE, eval = TRUE}
test_result <- readRDS(paste0(output_dir, "data/HMC_convergence_control_STAT1_STAT2.rds"))
```

The p-value from the convergence analysis is:

```{r, eval = TRUE}
test_result$p_value
```

A small p-value suggests that `STAT1` and `STAT2` induce similar effects, relative to the control group. The null hypothesis is that `STAT1` and `STAT2` perturb different sets of genes.

To further investigate which features contribute to the observed similarity or difference, we extract the consistently selected features:

```{r, eval = TRUE}
collect_active_features(test_result)
```

There is also a visualization function to examine the candidate convergence gene (`first_beta`) and selected convergence gene (`final_beta`, `second_pc` and `proj_direction`). Among these, the `proj_direction vector` is used to compute the test statistic within each fold. By visualizing and comparing these vectors across folds, we can assess the consistency of gene selection and projection direction.

```{r}
combined_df <- visualize_convergence_top_genes(
  fold_data = test_result$fold_data[1:2],
  top_n = 20
)
```

## 2. Performance when No Convergence

To illustrate how the method behaves under a null-like scenario, we simulate an artificial example where one of the treatment groups (treatment2) is nearly identical to the control group, except for a small perturbation in a few features.

Specifically, we:

-   Split the control group into two parts: `control_subset` and `psudo_control_subset`

-   Perturb the first 5 features of `psudo_control_subset` by multiplying them by 10

-   Compare `STAT1` and this perturbed subset of control, using `control_subset` as the baseline

```{r, echo = TRUE, eval = TRUE}
control_subset <- control[1:100, ]
psudo_control_subset <- control[101:200,]
psudo_control_subset[, 1:5] <- 10 * psudo_control_subset[, 1:5] 
psudo_control_subset[1:3, 1:5]
```

```{r, echo = TRUE, eval = FALSE}
set.seed(123)
test_result_null <- convergence_testing(
  control = control_subset,
  treatment1 = STAT1,
  treatment2 = psudo_control_subset,
  pca_method = "dense_pca",
  classifier_method = "lasso",
  lambda_type = "lambda.min",
  n_folds = 5,
  verbose = TRUE
)
```

```{r, echo = FALSE, eval = FALSE}
saveRDS(test_result_null, file = paste0(output_dir, "data/HMC_convergence_control_STAT1_pseudonull.rds"))
```

```{r, echo = FALSE, eval = TRUE}
test_result_null <- readRDS(paste0(output_dir, "data/HMC_convergence_control_STAT1_pseudonull.rds"))
```

Although `psudo_control_subset` differs from the control group (due to the artificial perturbation), the way it differs is not aligned with how `STAT1` differs from the control. As a result, our method detects this mismatch and returns a large p-value, suggesting that `STAT1` and the pseudo-control are **not** similar in their effects.

```{r}
test_result_null$p_value
collect_active_features(test_result_null)
```


## 3. Perturbation Convergence on Gene Modules

So far, we have demonstrated how to assess perturbation convergence by evaluating each measured gene individually. This gene-level strategy often identifies multiple genes with the strongest signals. As an alternative, one can first group all measured genes into modules and then assess convergence at the module level.

For this tutorial, we provide a pre-calculated gene module vector, which you can download from our GitHub repository:

```{r, echo = TRUE, eval = FALSE}
### download
url <- "https://raw.githubusercontent.com/terrytianyuzhang/HMC/main/docs/data/gene_clustering_for_HMC.rds"
dest_file <- paste0(my_local_path, "gene_clustering_for_HMC.rds")
download.file(url, dest_file, mode = "wb") 

### load it
clustering <- readRDS(dest_file)
```

```{r, echo = FALSE, eval = TRUE}
clustering <- readRDS(paste0(output_dir, "data/gene_clustering_for_HMC.rds"))
```

Format the grouping vector (in total, we have 41 modules).
```{r}
grouping_vector <- clustering$cluster_index
names(grouping_vector) <- clustering$gene_name
head(grouping_vector)
tail(grouping_vector)
```

```{r, eval = FALSE}
set.seed(123)
gp_lasso_test_result <- convergence_testing(
  control = control[1:300,], #group lasso is slower than lasso so I sub-sampled
  treatment1 = STAT1,
  treatment2 = STAT2,
  pca_method = "dense_pca",
  ##use group_lasso instead of the default lasso
  classifier_method = "group_lasso", 
  lambda_type = "lambda.min",
  n_folds = 5,
  ## genes with the same grouping number belong to the same module
  group = grouping_vector,
  verbose = TRUE
)
```

```{r, echo = FALSE}
gp_lasso_test_result = readRDS(paste0(output_dir, "/data/HMC_convergence_control_STAT1_STAT2_grLasso.rds"))
```

```{r}
gp_lasso_test_result$p_value
collect_active_features(gp_lasso_test_result, 
                        group = grouping_vector, 
                        # Modules with >= 5 active genes are considered active
                        group_threshold = 5) 
```

We conclude that STAT1 and STAT2 induce similar effects on the genes in modules 26 and 31. Notably, the genes in module 31 are mostly associated with immune function.

```{r}
clustering[cluster_index == 31,]$gene_name
```

### Remark.

How to establish the modules effectively is a whole different question, which still attracts a lot of research interests. Fortunately,  `convergence_testing()` is flexible and can take any method's clustering result as its input---simply replace `grouping_vector` with your own, as long as it follows the correct named format. The module list provided in this tutorial partially followed the pipeline proposed by [CS_CORE](https://changsubiostats.github.io/CS-CORE/articles/CSCORE.html)


## 4. Normality of the Test Statistics

The test statistic returned by `convergence_testing()` is standard Gaussian under the null hypothesis of no convergence. We conduct a simple simulation study to empirically verify this behavior.

```{r}
library(HMC)
library(MASS)
library(ggplot2)

set.seed(123)
simulate_null_data <- function(n_samples=500, n_features=100) {
  mu_con <- rep(0, n_features)
  mu_tr1 <- c(rep(5, 5), rep(0, n_features - 5))
  mu_tr2 <- c(rep(0, 5),  rep(0, n_features - 10), rep(5, 5))
  
  ratio <- 3
  control <- mvrnorm(n_samples*ratio, mu_con, diag(c(1:5, rep(1, n_features - 5))))
  treatment1 <- mvrnorm(n_samples, mu_tr1, diag(n_features))
  treatment2 <- mvrnorm(n_samples, mu_tr2, diag(n_features))
  
  rownames(control) <- paste("Sample", 1:(n_samples*ratio), sep = "_")
  rownames(treatment1) <- paste("Sample", 1:n_samples, sep = "_")
  rownames(treatment2) <- paste("Sample", 1:n_samples, sep = "_")
  
  colnames(control) <- paste("Feature", 1:n_features, sep = "_")
  colnames(treatment1) <- paste("Feature", 1:n_features, sep = "_")
  colnames(treatment2) <- paste("Feature", 1:n_features, sep = "_")
  
  return(list(control = control, treatment1 = treatment1, treatment2 = treatment2))
}

# Simulation settings
n_simulations <- 100
n_samples <- 50
n_features <- 200
n_folds <- 5

test_statistics <- numeric(n_simulations)
```

```{r, eval = FALSE}
# Run simulations
for (i in 1:n_simulations) {
  set.seed(i)
  data <- simulate_null_data(n_samples = n_samples, n_features = n_features)
  result <- convergence_testing(data$control, data$treatment1, data$treatment2,
                                lambda_type = 'lambda.min', n_folds = n_folds)
  test_statistics[i] <- result$test_statistic
}
```

```{r, echo = FALSE, eval = FALSE}
saveRDS(test_statistics, file = paste0(output_dir, "data/HMC_convergence_normality.rds"))
```

```{r, echo = FALSE, eval = TRUE}
test_statistics <- readRDS(paste0(output_dir, "data/HMC_convergence_normality.rds"))
```

```{r}
# Histogram of test statistics

ggplot(data.frame(test_statistics), aes(x = test_statistics)) +
  geom_histogram(binwidth = 0.1, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Test Statistics under Null", x = "Test Statistic", y = "Count") +
  theme_minimal()
```

```{r}
# Q-Q Plot using ggplot2
qq_df <- data.frame(
  sample_quantiles = sort(test_statistics),
  theoretical_quantiles = qnorm(ppoints(n_simulations))
)

ggplot(qq_df, aes(x = theoretical_quantiles, y = sample_quantiles)) +
  geom_point(color = "steelblue", alpha = 0.8) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Q-Q Plot of Test Statistics", x = "Theoretical Quantiles (Normal)", y = "Sample Quantiles") +
  theme_minimal()
```

<!-- Next, we apply the test using a subset of genes corresponding to a specific cluster. We load the gene grouping file and isolate genes from cluster 31: -->

<!-- ```{r, echo = FALSE, eval = TRUE} -->

<!-- clustering <- readRDS(paste0(output_dir, "/data/gene_clustering_for_HMC.rds")) -->

<!-- gene_to_keep <- which(clustering$cluster_index == 31) -->

<!-- control_subset <- control[, ..gene_to_keep] -->

<!-- perturbed_subset <- perturbed[, ..gene_to_keep] -->

<!-- ``` -->

<!-- ```{r, echo = TRUE, eval = FALSE} -->

<!-- clustering <- readRDS("gene_clustering_for_HMC.rds") -->

<!-- gene_to_keep <- which(clustering$cluster_index == 31) -->

<!-- control_subset <- control[, ..gene_to_keep] -->

<!-- perturbed_subset <- perturbed[, ..gene_to_keep] -->

<!-- ``` -->
