---
title: "Performing Convergence Anlysis Using HMC"
format: html
editor: visual
---

This tutorial demonstrates how to perform convergence analysis using the `HMC` package. We will use a sample dataset, which can be downloaded using the following Bash command:

```{bash, eval = FALSE}
wget https://github.com/terrytianyuzhang/HMC/blob/main/docs/data/tiny_cleary_for_HMC.rds
```

To get started, install the `HMC` package from GitHub using the following command:

```{r, eval = FALSE}
devtools::install_github("terrytianyuzhang/HMC/HMC_package")
```

Next, load the required packages and data, then extract the control and treatment groups for the convergence analysis. We will use `STAT1` and `STAT2`. Based on their names, we expect them to induce similar effects when perturbed.

```{r, echo = TRUE, eval=FALSE}
library(data.table)
library(HMC)

# Load data
residual_subset <- readRDS("tiny_cleary_for_HMC.rds")
clustering <- readRDS("tiny_cleary_for_HMC.rds")

# Subset to control and treatment groups, then remove guide labels
control <- residual_subset[
  Guides_collapsed_by_gene == "non-targeting",
  !"Guides_collapsed_by_gene"
]
STAT1 <- residual_subset[
  Guides_collapsed_by_gene == "STAT1",
  !"Guides_collapsed_by_gene"
]

STAT2 <- residual_subset[
  Guides_collapsed_by_gene == "STAT2",
  !"Guides_collapsed_by_gene"
]
```

```{r, echo = FALSE, eval= TRUE}
library(data.table)
library(HMC)

# Load data
output_dir <- "~/Documents/HMC/docs/"
residual_subset <- readRDS(paste0(output_dir, "/data/tiny_cleary_for_HMC.rds"))
clustering <- readRDS(paste0(output_dir, "/data/gene_clustering_for_HMC.rds"))

# Subset to control and treatment groups, then remove guide labels
control <- residual_subset[
  Guides_collapsed_by_gene == "non-targeting",
  !"Guides_collapsed_by_gene"
]
STAT1 <- residual_subset[
  Guides_collapsed_by_gene == "STAT1",
  !"Guides_collapsed_by_gene"
]

STAT2 <- residual_subset[
  Guides_collapsed_by_gene == "STAT2",
  !"Guides_collapsed_by_gene"
]
```

First 3 cells and first 5 genes from the control group:

```{r, eval=TRUE}
control[1:3, 1:5]
```

First 3 cells and first 5 genes from the STAT1 group:

```{r, eval = TRUE}
STAT1[1:3, 1:5]
```

Now we perform the convergence analysis using the `convergence_testing()` function.

We compare `STAT1` and `STAT2`, using control group as the baseline.

```{r, echo = TRUE, eval = FALSE}
set.seed(123)
test_result <- convergence_testing(
  control = control,
  treatment1 = STAT1,
  treatment2 = STAT2,
  pca_method = "dense_pca",
  classifier_method = "lasso",
  lambda_type = "lambda.min",
  n_folds = 5,
  verbose = TRUE
)
```

```{r, echo = FALSE, eval = FALSE}
saveRDS(test_result, file = paste0(output_dir, "data/HMC_convergence_control_STAT1_STAT2.rds"))
```

```{r, echo = FALSE, eval = TRUE}
test_result <- readRDS(paste0(output_dir, "data/HMC_convergence_control_STAT1_STAT2.rds"))
```

The p-value from the convergence analysis is:

```{r, eval = TRUE}
test_result$p_value
```

A small p-value suggests that `STAT1` and `STAT2` induce similar effects, relative to the control group. The null hypothesis is that `STAT1` and `STAT2` perturb different sets of genes.

To further investigate which features contribute to the observed similarity or difference, we extract the consistently selected features:

```{r, eval = TRUE}
collect_active_features(test_result)
```

To illustrate how the method behaves under a null-like scenario, we simulate an artificial example where one of the treatment groups (treatment2) is nearly identical to the control group, except for a small perturbation in a few features.

Specifically, we:

-   Split the control group into two parts: `control_subset` and `psudo_control_subset`

-   Perturb the first 5 features of `psudo_control_subset` by multiplying them by 10

-   Compare `STAT1` and this perturbed subset of control, using `control_subset` as the baseline

```{r, echo = TRUE, eval = TRUE}
control_subset <- control[1:100, ]
psudo_control_subset <- control[101:200,]
psudo_control_subset[, 1:5] <- 10 * psudo_control_subset[, 1:5] 
psudo_control_subset[1:3, 1:5]
```

```{r, echo = TRUE, eval = FALSE}
set.seed(123)
test_result_null <- convergence_testing(
  control = control_subset,
  treatment1 = STAT1,
  treatment2 = psudo_control_subset,
  pca_method = "dense_pca",
  classifier_method = "lasso",
  lambda_type = "lambda.min",
  n_folds = 5,
  verbose = TRUE
)
```

```{r, echo = FALSE, eval = FALSE}
saveRDS(test_result_null, file = paste0(output_dir, "data/HMC_convergence_control_STAT1_pseudonull.rds"))
```

```{r, echo = FALSE, eval = TRUE}
test_result_null <- readRDS(paste0(output_dir, "data/HMC_convergence_control_STAT1_pseudonull.rds"))
```

Although `psudo_control_subset` differs from the control group (due to the artificial perturbation), the way it differs is not aligned with how `STAT1` differs from the control. As a result, our method detects this mismatch and returns a large p-value, suggesting that `STAT1` and the pseudo-control are **not** similar in their effects.

```{r}
test_result_null$p_value
collect_active_features(test_result_null)
```

<!-- Next, we apply the test using a subset of genes corresponding to a specific cluster. We load the gene grouping file and isolate genes from cluster 31: -->

<!-- ```{r, echo = FALSE, eval = TRUE} -->

<!-- clustering <- readRDS(paste0(output_dir, "/data/gene_clustering_for_HMC.rds")) -->

<!-- gene_to_keep <- which(clustering$cluster_index == 31) -->

<!-- control_subset <- control[, ..gene_to_keep] -->

<!-- perturbed_subset <- perturbed[, ..gene_to_keep] -->

<!-- ``` -->

<!-- ```{r, echo = TRUE, eval = FALSE} -->

<!-- clustering <- readRDS("gene_clustering_for_HMC.rds") -->

<!-- gene_to_keep <- which(clustering$cluster_index == 31) -->

<!-- control_subset <- control[, ..gene_to_keep] -->

<!-- perturbed_subset <- perturbed[, ..gene_to_keep] -->

<!-- ``` -->
